---
title: "Reptiles in the ACT"
subtitle: Occurrences across space and time
author: "Shandiya Balasubramaniam"
output: 
  html_document:
    code_folding: "hide"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r libraries, message = FALSE, warning = FALSE}

library(here)
library(galah)
library(skimr)
library(dplyr)
library(lubridate)
library(tidyr)
library(stringr)
library(ggplot2)
library(leaflet)
library(leaflet.extras)
library(reactable)
library(scico)
library(hrbrthemes)

```


```{r get-data, include = FALSE, message = FALSE, echo = FALSE}

# email in .Renviron
galah_config(email = Sys.getenv('EMAIL'), atlas = "Australia")

my_col <- select_columns("decimalLatitude",
                         "decimalLongitude",
                         "eventDate",
                         "scientificName", 
                         "genus",
                         "family",
                         "order",
                         "dataResourceName", 
                         "basisOfRecord", 
                         "cl22",
                         "cl10902")
# download data
reptilia <- ala_occurrences(
  taxa = select_taxa("Reptilia"),
  filters = select_filters(stateProvince = "Australian Capital Territory"),
  columns = my_col,
  mint_doi = TRUE)

citation_text <- ala_citation(reptilia)

# save to avoid repeated downloads
saveRDS(reptilia, here("data", "reptilia"))

```


```{r tidying, message = FALSE}

reptilia <- readRDS(here("data", "reptilia"))

reptilia_tidy <- reptilia %>% 
  filter(cl22 == "Australian Capital Territory") %>% 
  rename(state = cl22, 
         forestType = cl10902) %>% 
  mutate(forestType = na_if(forestType, ""),
         family = na_if(family, ""),
         basisOfRecord = tolower(basisOfRecord),
         basisOfRecord = str_replace(basisOfRecord, "_", " "),
         eventDate = as_datetime(eventDate, tz = "Australia/Sydney", format = NULL),
         eventDate = as_date(eventDate, tz = NULL),
         month = month(eventDate))

```

## Exploratory data analysis {.tabset .tabset-fade .tabset-pills}

### Summary statistics

Summary statistics for fields in the dataset of reptile occurrences in the ACT

```{r skim, fig.fullwidth = TRUE}

skim(reptilia_tidy)

```


### Occurrence map

Locations of reptiles in the ACT based on occurrence records of reptiles from the ALA. Interactive layers in the top right corner allow records to be differentiated based on the type of record (ALA id: basisOfRecord), forest type (ALA id: cl10902), and taxonomic family (ALA id: family) associated with each record.

```{r leaflet, echo = FALSE, layout = "l-page"}

# colour palettes
record_pal <- colorFactor(
  scico(5, palette = 'lajolla'), 
  reptilia_tidy$basisOfRecord
)


forest_pal <- colorFactor(
  scico(11, palette = 'romaO'),
  reptilia_tidy$forestType
)

family_pal <- colorFactor(
  scico(12, palette = 'batlow'),
  reptilia_tidy$family
)

leaflet(reptilia_tidy) %>% 
  # different basemaps 
  addProviderTiles(providers$CartoDB.Positron, group = "Positron") %>% 
  addProviderTiles(providers$Stamen.Toner, group = "Toner") %>%
  addProviderTiles(providers$Stamen.Terrain, group = "Terrain") %>% 
  # all records
  addCircleMarkers(
    lng = ~decimalLongitude,
    lat = ~decimalLatitude,
    radius = 3,
    stroke = FALSE,
    color = "#41507B",
    fillOpacity = 0.5,
    group = "All records",
    popup = paste0(reptilia_tidy$scientificName)
  ) %>% 
  # colours points by basis of record
  addCircleMarkers(
    lng = ~decimalLongitude,
    lat = ~decimalLatitude,
    radius = 3,
    stroke = FALSE,
    color = ~record_pal(basisOfRecord),
    fillOpacity = 0.5,
    group = "Record basis",
    popup = paste0(reptilia_tidy$scientificName, "<br/>", reptilia_tidy$basisOfRecord)
  ) %>%
  # legend for basis of record
  addLegend(
    "bottomright",
    pal = record_pal,
    values = ~ basisOfRecord,
    title = "Basis of Record",
    opacity = 1,
    group = "Record basis"
  ) %>% 
  # colours points by forest type
  addCircleMarkers(
    lng = ~decimalLongitude,
    lat = ~decimalLatitude,
    radius = 3,
    stroke = FALSE, 
    color = ~forest_pal(forestType),
    fillOpacity = 0.5,
    group = "Forest type",
    popup = paste0(reptilia_tidy$scientificName, "<br/>", reptilia_tidy$forestType)
  ) %>% 
  # legend for forest type
  addLegend(
    "bottomright",
    pal = forest_pal,
    values = ~ forestType,
    title = "Forest Type",
    opacity = 1,
    group = "Forest type"
  ) %>% 
  # colours points by family
  addCircleMarkers(
    lng = ~decimalLongitude,
    lat = ~decimalLatitude,
    radius = 3,
    stroke = FALSE,
    color = ~family_pal(family),
    fillOpacity = 0.5,
    group = "Family",
    popup = paste0(reptilia_tidy$scientificName, "<br/>", reptilia_tidy$family)
  ) %>%
  # legend for family
  addLegend(
    "bottomright",
    pal = family_pal,
    values = ~family,
    title = "Family",
    opacity = 1,
    group = "Family"
  ) %>% 
  addLayersControl(
    baseGroups = c("Positron", "Toner", "Terrain"),
    overlayGroups = c("All records", "Record basis", "Forest type", "Family"), 
    options = layersControlOptions(collapsed = FALSE)) %>% 
  hideGroup(c("Record basis", "Forest type", "Family")) %>% 
  addFullscreenControl()

```
  

### Occurrence records

Occurrence records for reptiles in the ACT. Columns may be sorted by clicking on column names, or filtered using the search bars under each column header. A global search bar in the top right corner enables searching the whole table.  

```{r table-all, fig.fullwidth = TRUE}

# remove cols not of interest in a table
reptilia_subset <- reptilia_tidy %>% 
  select(eventDate:basisOfRecord, forestType)

reactable(
  reptilia_subset,
  columns = list(
    eventDate = colDef(name = "Date"),
    scientificName = colDef(name = "Species"),
    genus = colDef(name = "Genus"),
    family = colDef(name = "Family"),
    order = colDef(name = "Order"),
    dataResourceName = colDef(name = "Source"),
    basisOfRecord = colDef(name = "Record type"),
    forestType = colDef(name = "Forest type")
  ),
  showSortable = TRUE,
  filterable = TRUE,
  searchable = TRUE,
  striped = TRUE
)
```

## {-}


## Occurrence records over time {.tabset .tabset-fade .tabset-pills}

### All records   

Reptile occurrence counts between `r year(min(reptilia_tidy$eventDate, na.rm = TRUE))` and `r year(max(reptilia_tidy$eventDate, na.rm = TRUE))` in the ACT

```{r all-counts, message = FALSE, fig.fullwidth = TRUE}

reptilia_tidy %>%
  filter(!is.na(eventDate)) %>% 
  group_by(eventDate) %>%
  summarise(count = n()) %>%
  ggplot() +
  geom_line(aes(x = eventDate, y = count), 
            color = "#41507B") +
  labs(x = "",
       y = "Count") +
  theme_ipsum()

```

### Records by family   

Reptile occurrence counts by taxonomic family between `r year(min(reptilia_tidy$eventDate, na.rm = TRUE))` and `r year(max(reptilia_tidy$eventDate, na.rm = TRUE))` in the ACT 

```{r facet-counts, message = FALSE, fig.fullwidth = TRUE}

reptilia_tidy %>%
  filter(!is.na(eventDate)) %>% 
  group_by(year(eventDate)) %>%
  summarise(count = n(), family = family) %>%
  ggplot() +
  geom_line(aes(x = `year(eventDate)`, y = count), 
            color = "#41507B") +
  facet_wrap(~family) +
  labs(x = "Date",
       y = "") +
  theme_ipsum(axis_text_size = 9)

```

### Records by data resource   

Reptile occurrence counts based on input source (data resource name) between `r year(min(reptilia_tidy$eventDate, na.rm = TRUE))` and `r year(max(reptilia_tidy$eventDate, na.rm = TRUE))` in the ACT 

```{r data-resource, message = FALSE, fig.fullwidth = TRUE}

reptilia_tidy %>%
  filter(!is.na(eventDate)) %>% 
  group_by(year(eventDate)) %>%
  summarise(count = n(), source = dataResourceName) %>%
  ggplot() +
  geom_line(aes(x = `year(eventDate)`, y = count), 
            color = "#41507B") +
  facet_wrap(~source) +
  labs(x = "Date",
       y = "") +
  theme_ipsum(
    strip_text_size = 9, 
    axis_text_size = 9)

```

### Seasonal differences in counts  

Occurrence counts of individuals by family, across the different months of the year. Although the number of counts varies widely among families, the patterns across months are similar. NOTE: counts have not been transformed to account for differences in the number of species in each family.  

```{r polar-plots, message = FALSE, fig.fullwidth = TRUE}

# wrangling
reptilia_polar <- reptilia_tidy %>%
  filter(!is.na(month)) %>%                                # remove records w/o date
  group_by(family, month) %>%
  summarise(occ = n()) %>%                                 # counts by family and month
  pivot_wider(names_from = month, values_from = occ) %>%   # for filtering out low counts by family
  mutate_if(is.numeric, replace_na, 0) %>%                 # replace NAs with 0s so sum() works   
  mutate(total_by_family = sum(c_across(1:12))) %>%
  filter(total_by_family > 50) %>%                         # remove families with <50 occ total
  mutate(`0` = `12`) %>%                                   # adds dummy month column for nicer plotting
  select(-total_by_family) %>%
  pivot_longer(!family, names_to = "month", values_to = "occ") %>%
  mutate(month = as.numeric(month))                        # allows plotting as continuous variable

# temp solution to enable free scales in polar plot
cp <- coord_polar()
cp$is_free <- function() TRUE

# polar plot facetted by family
ggplot(reptilia_polar) +
  geom_line(aes(x = month, y = occ), 
            colour = "#41507B", lwd = 1) +
  scale_x_continuous(breaks = 1:12) + 
  cp + 
  facet_wrap(~family, scales = "free_y") +
  theme_ipsum(axis_text_size = 9) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(), 
        panel.grid.minor = element_blank())

```

## {-}

### Citation  
`r citation_text`


### Session Info  
```{r session-info}
sessionInfo()
```

